---
title: "Homework Assignment 1"
author: "Your Names Here"
date: 'Assigned: Mar 11, 2021, Due Mar 21, 2021 11:59PM'
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

## HW-1 Instructions

> Our class emphasizes clear communication of data analysis results to non-technical audiences. I expect your HTML output documents to be readable and well formatted. I expect you to put ample comments in your R code to make the code understandable. Along with accuracy of results, I am looking for quality of presentation as well. To complete this assignment, follow these steps:

1. Download the `HW1.Rmd` file from Canvas.

2. Open `HW1.Rmd` in RStudio.

3. Replace the "Your Name Here" text in the `author:` field with names of the students in HW team - up to two students can work together on HWs. Note that each student needs to submit his/her assignment separately.

4. Supply your solutions to the homework by editing `HW1.Rmd`.

5. When you have completed the homework and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, you should then submit both the RMD file and its html output.

6. Run your code in the Console and Knit HTML frequently to check for errors.

7. You may find it easier to solve a problem by interacting only with the Console at first. 

8. Note - you can insert an R block for writing your R code by either clicking the **insert** button or pressing Control-Alt-I. You can knit an RMD file into its HTML output by clicking the **Knit** button or pressing "Control-Shift-K".


We will continue working with the nycflights dataset we looked at last class. **Please be sure to keep the data file in the same directory as the RMD file.**

First - lets read the data into an object named `nyc`.

```{r}
nyc <- read.csv("nycflights.csv")
#Lets calculate how many rows there are
nrow(nyc)
hist(nyc$arr_delay)
```

Number of  rows in my dataset is :`r nrow(nyc)`

I now would like you to answer the following - all in their own separate R code blocks.

### Data Exploration

Let's first do some simple exploration of this data. I would like for you to answer each of the questions below with an **inline R code** integrated into the text and not in a separate R code block.

- How many airlines are there? (Hint: `levels` and `length` can be useful here)

```{r, echo=FALSE, include=FALSE}
#We will calculate how many airlines there are 
#First- lets see the structure of the data
str(nyc)

#convert carrier to a factor
nyc$carrier <- as.factor(nyc$carrier)

length(levels(nyc$carrier))
```

There are `r length(levels(nyc$carrier))` different airlines in our dataset.

- How many flights there were by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)

```{r}
#LOgical indexing 

nycoo <- nyc[  nyc$carrier =="OO","carrier"   ]
length(nycoo)


#Simple logical Indexing 
nyc[c(1,2,3:10),]

nrow(nyc[nyc$month ==5 & nyc$des =="DTW",])

delayairport <- nyc[nyc$arr_delay >= 120 ,"dest"]
sort(table(delayairport),decreasing = TRUE) [1:5]

max(nyc$distance)
mean(nyc$arr_delay,na.rm =TRUE)
```


- How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)

min(nyc$arr_time, na.rm =T)

- How many flights where there by United Airlines (code: UA) on Jan 12th 2013?
```{r}
nrow(nyc[nyc$carrier =="UA" & nyc$month ==1 & nyc$year== 2013 & nyc$day ==12, ])

```



### Arrival Delay

Lets focus on Arrival Delay.

- What was the average arrival delay for all airports and all airlines combined in Jan 2013? 
```{r}
mean(nyc[nyc$month==1 & nyc$year==2013, "arr_delay"],na.rm=T)
```


- Whats was the median arrival delay for all airports and all airlines combined in Jan 2013?
```{r}
median(nyc[nyc$month==1 & nyc$year==2013, "arr_delay"],na.rm=T)
```


Based on your answers to the two questions above, what can you say about the distribution of arrival delays? Provide your answer in a text paragraph form.

### Airline Performance

Lets see if all airlines are equally terrible as far as flight arrival delays are concerned. For this question you will have to make sure that airline column is coded as a factor.

- Calculate average arrival delays by airline (Hint: look up the command `tapply`)

```{r}
arrivaldelay <- round(tapply( nyc$arr_delay, nyc$carrier, mean, na.rm=TRUE),2)
arrivaldelay
barplot(sort(arrivaldelay, decreasing =TRUE), cex.names =0.6)
```



- Draw a Bar Plot of Average Arrival Delays for all the Airlines (Hint: command for making a Bar Plot is simply `barplot`)

```{r}
barplot(sort(arrivaldelay, decreasing =TRUE), cex.names =0.6)

max(arrivaldelay)
names(arrivaldelay[arrivaldelay==max(arrivaldelay)])
length(arrivaldelay[arrivaldelay < 0])

```


- Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay? Provide answer to this question in a text paragraph using **inline R code**.

The airline with highest average arrival delay is `r names(arrivaldelay[arrivaldelay==max(arrivaldelay)])`. The airlines with  the smallest arrival delay is `r names(arrivaldelay[arrivaldelay==min(arrivaldelay)])`.

These are `r length(arrivaldelay[arrivaldelay < 0])` airlines with negative  average delay : `r names(arrivaldelay[arrivaldelay < 0])`.

That's it. Once you are done, make sure everything works and knits well and then you can uplaod the RMD flile and the html output to Canvas.


**Have Fun!**

Sanjeev
